---
title: "BotEx: Using LLMs as Experimental Participants in oTree"
subtitle: "" # we use this for the venue
author: Joachim Gassen
institute: Humboldt-Universität zu Berlin
date: today
date-format: "MMMM D, YYYY"
format: 
  beamer: 
    latex_engine: xelatex # pdflatex creates rastered fonts
    slide_level: 3
    
classoption: "aspectratio=169"

header-includes:
- \usepackage{booktabs} 
- \usepackage{tabularx}
- \usepackage{multirow,makecell}
- \usepackage{array}
- \renewcommand\tabularxcolumn[1]{m{#1}}
- \usepackage{makecell}
- \usepackage{colortbl}
- \usepackage{adjustbox}
- \usepackage{tikz}
- \usepackage{siunitx}
- \usepackage{tabu}
- \usetikzlibrary{arrows,arrows.meta,calc,positioning,matrix}
- \input{materials/beamer_theme_trr266_16x9.sty}
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(
  cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE
)

suppressMessages({
  library(tidyverse)
  library(ggbeeswarm)
  library(kableExtra)
  library(modelsummary)
  library(showtext)
	library(ggwordcloud)
	library(tidytext)
	
  devtools::source_url(
    "https://raw.githubusercontent.com/trr266/treat/main/code/R/theme_trr.R"
  )
})

options(modelsummary_format_numeric_latex = "plain")

rounds <- read_csv("../data/generated/rounds.csv", show_col_types = FALSE)
participants <- read_csv("../data/generated/participants.csv", show_col_types = FALSE)
rounds$pct_returned <- rounds$sent_back_amount/(3*rounds$sent_amount)

# Needs \usepackage{adjustbox} in preamble
fit_gt_table_to_slide <- function(x, scale = 1) {
	if (scale == 1) repstr <- "\\\\resizebox{\\\\columnwidth}{!}{\\\\begin{tabular}"
	else repstr <- paste0("\\\\resizebox{", scale, "\\\\columnwidth}{!}{\\\\begin{tabular}")
  x <- sub("\\\\begin\\{longtable\\}", repstr, x)
  x <- sub("\\\\end\\{longtable\\}", "\\\\end{tabular}}", x)
  x
}

fit_gt_table_to_slide_long <- function(x, scale = 1) {
	if (scale == 1) repstr <- "\\\\resizebox*{!}{\\\\textheight}{\\\\begin{tabular}"
	else repstr <- paste0("\\\\resizebox*{!}{", scale, "\\\\textheight}{\\\\begin{tabular}")
  x <- sub("\\\\begin\\{longtable\\}", repstr, x)
  x <- sub("\\\\end\\{longtable\\}", "\\\\end{tabular}}", x)
  x
}

center_table <- function(x) {
	x[1] <- paste0("\\begin{center}", x[1], "\\end{center}")
	x
}
```

```{r InstallTRRFonts}
trr266_fonts_available <- all(c("Cambria", "Calibri Light") %in% font_families())

if (!trr266_fonts_available) {
  # On Mac with MS Office installation you need to add the Office
  # Fonts to the font path. The following depends on your 
  # installation but _should_ work in most cases
  
  office_font_path <- "/Applications/Microsoft Word.app/Contents/Resources/DFonts"
  if (Sys.info()["sysname"] == "Darwin") {
    if (dir.exists(office_font_path)) font_paths(office_font_path)
    else stop("MS Office font path not found")
  }
  
  rv <- tryCatch({
    font_add(family = "Cambria", regular = "Cambria.ttc")
    font_add(family = "Calibri Light", regular = "calibril.ttf")
  }, error = function(e) {
    message(sprintf("Failed to install TRR 266 fonts: %s", e))
    invisible(font_families())        
  })
  
  trr266_fonts_available <- all(c("Cambria", "Calibri Light") %in% rv)
  if (trr266_fonts_available) message("Successfully installed TRR 266 fonts") 
}
```

## Motivation

1. Designing experiments is hard and small differences in experimental materials can have significant impact on experimental findings. Wouldn't it be great to have a "cheap" way to pretest experimental designs without bothering humans?

2. Understanding how LLMs act in behavioral experiments - on their own and interacting with humans - is an emergent and rapidly evolving research field. Providing an infrastructure to run such experiments should be beneficial for the profession.

3. Applied experimental work often uses context framing in their experimental designs without explicitly hypothesizing and assessing its effect on findings. LLM based experiments might inform priors in that regard.


## Traditional oTree Setup

\begin{center}
\includegraphics[height =0.7\textheight]{"materials/otree.jpeg"}
\end{center}

## oTree + Alter Ego

\begin{center}
\includegraphics[height =0.7\textheight]{"materials/otree_alterego.jpeg"} \\
Engel, Grossmann and Ockenfels (2024, SSRN)
\end{center}

## oTree + BotEx

\begin{center}
\includegraphics[height =0.7\textheight]{"materials/otree_botex.jpeg"}
\end{center}


## A First Example

- As a first sniff test, I study whether an investor/manager framing in a classic multi-round investment trust game (Berg, Dickhaut and McCabe, Games and Econ Behav 1995) affects the observed outcome.

- Prior work has shown that the investment trust game is subject to context framing effects (e.g., Al-Ubaydli, Houser,
Nye, Paganelli, and Pan, PLOS One 2013; Cronk and Wasielewski, J of Evolutionary Psychology 2008)

- Can similar framing effects also be observed for LLMs?


## The Neutral Frame

\begin{center}
\includegraphics[height =0.85\textheight]{"materials/otree_inst_neutral.jpeg"}
\end{center}


## The Investor/Manager Frame

\begin{center}
\includegraphics[height =0.85\textheight]{"materials/otree_inst_framed.jpeg"}
\end{center}


## Sample Composition

|                             |   Sender | Receiver |
|:----------------------------|---------:|---------:|
| Neutral Framing             |       15 |       15 |
| Investor/Management Framing |       15 |       15 |
| Total                       |       30 |       30 |


## Descriptive Statistics by Framing

```{r DescStatsTab}
datasummary(sent_amount + sent_back_amount + pct_returned ~ experiment*(N + mean + sd + median), data = rounds, )

```

## Sent Amount by Framing

```{r SentAmountFig, fig.width=8, fig.height=3.5, fig.align="center", fig.showtext = TRUE}
color_scale <- RColorBrewer::brewer.pal(3 ,"Set1")
names(color_scale) <- unique(rounds$experiment)
color_scale_labs <- c("Investor/Manager", "Neutral")

ggplot(rounds, aes(x = experiment, y = sent_amount, group = experiment, color = experiment)) +
  geom_boxplot() +
	geom_beeswarm(size = 0.25) +
  labs(x = "", y = "") +
	scale_x_discrete(labels = NULL, breaks = NULL) +
	scale_color_manual("", values = color_scale, labels = color_scale_labs) +
	theme_minimal() +
	theme(
		panel.grid.minor = element_blank()
	)

```

## Sent Amount by Round

```{r SentAmountRoundFig, fig.width=8, fig.height=3.5, fig.align="center", fig.showtext = TRUE}
ggplot(rounds, aes(x = as.factor(round), y = sent_amount, group = experiment, color = experiment)) +
	geom_beeswarm(size = 0.25, corral = "wrap") +
	stat_summary(
		aes(fill = experiment), color = NA, 
		geom = "ribbon", fun = "mean", alpha=0.1,
		fun.min = function(x) mean(x) - 1.96 / (length(x) - 1) * sd(x), 
		fun.max = function(x) mean(x) + 1.96 / (length(x) - 1) * sd(x),
		show.legend = FALSE
	) + 
	labs(x = "Round", y = "", color = "") +
	scale_color_manual("", values = color_scale, labels = color_scale_labs) +
	scale_fill_manual("", values = color_scale, labels = color_scale_labs) +
	theme_minimal() +
	theme(
		panel.grid.major.x  = element_blank(),
		panel.grid.minor = element_blank()
	)
```

## Regression Results

```{r RegressionTab}
modelsummary(
	lm(sent_amount ~ round*experiment, data = rounds),
	stars = c(`***` = 0.01, `**` = 0.05, `*` = 0.1),
	gof_omit = "BIC|AIC|F|RMSE|Log|^R2$"
)
```

## Some verbal response

```{r ReasoningFig, fig.width=8, fig.height=3.5, fig.align="center", fig.showtext = TRUE}
vr <- rounds %>%
	mutate(full_reason = paste(sent_reason, sent_back_reason)) %>%
	unnest_tokens(word, full_reason) %>%
	anti_join(stop_words, by = "word") %>% 
	group_by(experiment, word) %>%
	filter(n() > 50) %>%
	summarise(count = n())

ggplot(vr, aes(label = word, size = count, color = experiment)) +
	geom_text_wordcloud() +
	scale_color_manual("", values = color_scale, labels = color_scale_labs) +
	facet_wrap(~experiment) +
	theme_minimal()

```



## Manipulation checks ;-)

```{r ManCheckTab}
df <- participants %>%
	mutate(
		comprehension_check = factor(ifelse(comprehension_check == 3, "Yes", "No"), c("Yes", "No")),
		manipulation_check =  factor(ifelse(manipulation_check == role_in_group, "Yes", "No"), c("Yes", "No")),
		human_check = factor(ifelse(human_check == 1, "Human", "Bot"), c("Human", "Bot"))
	)
datasummary(
	(`Answers comprehension question correctly` = comprehension_check) + 
		(`Remembers role` = manipulation_check) + 
		(`Characterizes as ...` = human_check) ~ experiment*N, 
	data = df)

```

## Summary and Next Steps 

It works (at least for a very simple albeit repeated game):

- Context framing as investor/manager setting increases willingness of the sender to send money, and leaves strategy of the receiver unaffected
- Costs: ~30 US-$ for 60 participants, duration: app. 4 hours (fully scriptable)

Next steps

- Make framework portable 
- Other games?
- Identify seminal accounting study to compare to its "neutral" alternative 
- Alternatively: replicate Liberman, Samuels and Ross (2014) study on "Community" vs. "Wallstreet" framing for prisoner's dilmma  (but: likelihood that results are in LLM training material...)
- Compare findings with results for human participants 
